Software Architectures to Integrate Workflow Engines in Science Gateways

### 2016-04-26 paper.tex
Author addition
Samir should be added?

\definecolor{ao(english)}{rgb}{0.0, 0.5, 0.0}
\newcommand{\todo}[2]{\pdfmargincomment[color=red,author=#1,open=true]{#2}}
\newcommand{\note}[2]{\pdfmargincomment[color=yellow,author=#1,open=true]{#2}}
\newcommand{\closednote}[4]{\pdfmargincomment[color=gray,author=#1-#3,open=true]{#1: #2 \textCR \textCR #3: #4}}
\newcommand{\answerednote}[4]{\pdfmargincomment[color=green,author=#1-#3,open=true]{#1: #2 \textCR \textCR #3: #4}}

@@

@@

Robustness is a highly desirable characteristic in production gateways. In this respect, the 
multiplicity of software components and interactions involved in architectures is
a challenge as it often leads to complex systems.
Architectures showing reduced system complexity will greatly
facilitate the implementation effort towards robust interactions
which, in turn, have a positive impact on characteristics such as
gateway predictability, transparence, general reliability and results
traceability and reproducibility; characteristics that are also key to
a good user experience. \closednote{Marc-Etienne}{I find that
  something is weak in this last paragraph. Robustness will help with
  transparency, sure, but it seems only a small facet.  Also, the text
  seem to timidly favor architectures with less
  components/interactions without much explanation... Complex systems
  can be robust, but at greater costs...  Here is one: "Architectures
  showing reduced system complexity will greatly facilitate the
  implementation effort towards robust interactions which, in turn,
  have a positive impact on characteristics such as gateway
  predictability, transparence, general reliability and results
  traceability and reproducibility; characteristics that are also key
  to a good user experience."}{Tristan}{Agreed, fixed as suggested.}

@@

Various science gateways have been developed, including frameworks
used as building blocks to assemble science gateways. The main
frameworks are Apache Airavata~\cite{marru2011apache}, the Catania
Science Gateway Framework~\cite{Ardizzone2012} and
WS-PGRADE/gUSE~\cite{Kacsuk2012}. Numerous science gateway instances
were built using such
frameworks~\cite{kacsuk2014science,ardizzone2012decide}. Other gateway
instances were also built independently from any framework, using
regular web development tools and libraries to access distributed
systems. CBRAIN and VIP are two such science gateways, from which we
gathered experience about the architectures described in this paper.

@@

\begin{figure*}
\centering
\hspace*{0.2\columnwidth}
\subfloat[Tight integration]{
\def\svgwidth{0.6\columnwidth}
\input{figures/tight.pdf_tex}
\label{archi:tight}
}
\hfill
\subfloat[Service invocation]{
\def\svgwidth{0.6\columnwidth}
\input{figures/service.pdf_tex}
\label{archi:service}
\hspace*{0.2\columnwidth}
}\\
\hspace*{0.2\columnwidth}
\subfloat[Sub-tasking]{
\def\svgwidth{0.6\columnwidth}
\input{figures/sub-task.pdf_tex}
\label{archi:sub-task}
}
\hfill
\subfloat[Pool]{
\def\svgwidth{0.6\columnwidth}
\input{figures/agent.pdf_tex}
\label{archi:agent}
\hspace*{0.2\columnwidth}
}\\
\subfloat[Nested workflows. Left: abstract model. Right: instantiation with service invocation.]{
\def\svgwidth{0.9\columnwidth}
\input{figures/nested-1.pdf_tex}
\def\svgwidth{0.9\columnwidth}
\input{figures/nested-2.pdf_tex}
\label{archi:nested}
}\\
\subfloat[Workflow import. Left: abstract model. Right: instantiation with service invocation.]{
\def\svgwidth{0.5\columnwidth}
\input{figures/import-1.pdf_tex}
\hfill
\def\svgwidth{0.5\columnwidth}
\input{figures/import-2.pdf_tex}
\label{archi:import}
}
\caption{Architectures}
\label{fig:architectures}
\end{figure*}
\begin{table*}
\centering
\footnotesize
\rowcolors{1}{white}{gray!25}
\begin{tabular}{ll}
  \textbf{Architecture} & \textbf{Systems} \\
  \hline
  Tight & \pbox{1.5\columnwidth}{
          Catania Science Gateway Framework~\cite{Ardizzone2012},
          Distributed application runtime environment (DARE~\cite{maddineni2012distributed}),
          DECIDE~\cite{ardizzone2012decide}$^{c,n}$, LONI Pipeline Environment$^n$~\cite{dinov2009efficient}
          }
  \\
  Service & \pbox{1.5\columnwidth}{
            Apache Airvata~\cite{marru2011apache}, e-BioInfra~\cite{shahand2015data}$^{w,n}$, HubZero with Pegasus~\cite{CPE:CPE3257}, MoSGrid~\cite{krüger2014mosgrid}$^w$, System in~\cite{wu2010accelerating}, Vine Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual Imaging Platform~\cite{GLAT-13}$^n$, WS-PGRADE/gUSE framework~\cite{Kacsuk2012},
            Science gateways in~\cite{kacsuk2014science}$^w$
            } \\
  Sub-task & CBRAIN and PSOM~\cite{GLAT-16}$^n$, CBRAIN and FSL$^n$\\
  Pool & SHIWA pool~\cite{ROGE-13}\\
  Nested & \pbox{1.5\columnwidth}{SHIWA Simulation Platform (Coarse-Grained Interoperability~\cite{terstyanszky2014enabling})$^w$, HubZero with Pegasus (via hierarchical workflows)~\cite{Deelman201517}, Tavaxy~\cite{Abouelhoda2012}.
           }\\
  Import & SHIWA Simulation Platform (Fine-Grained Interoperability)~\cite{plankensteiner-prodan-etal:2013}$^w$, Tavaxy~\cite{Abouelhoda2012}, system in~\cite{delaGarza2016}.
\end{tabular}
\caption{Classification of science gateways based on the architecture used to integrate workflow engines. $^c$: based on the Catania Science Gateway Framework. $^w$: based on WS-PGRADE/gUSE. $^n$: used for neuroimaging.}
\label{table:system-classification}
\end{table*}

@@

\item Workflow integration: consists in adding a new workflow to the
  system so that users can execute it. It is triggered by an
  administrator of the science gateway and it results in an interface,
  for instance a web form, where users can enter the parameters of the
  workflow to be executed. The interaction has two aspects:
  (\texttt{a$_1$}) the programs used in the workflow are installed on
  the infrastructure, which may require administrative privileges on
  the infrastructure; (\texttt{a$_2$}) the workflow is configured in
  the science gateway so that it becomes available to users. Note that
  integrating a workflow is not the same process as integrating a
  workflow \emph{engine}. \note{Marc-e}{automation?}
\item Task control: operations to manage tasks on the infrastructure,
  including: authentication, submission, monitoring, termination,
  deletion, etc. Controlling tasks requires to deal with the
  heterogeneous batch managers and meta-schedulers that might be
  available on the infrastructure. When the infrastructure is a grid
  or a cloud, it may for instance be achieved using libraries that
  implement standards such as SAGA (Simple API for Grid
  Applications~\cite{goodale2006saga}), DRMAA (Distributed Resource
  Management Application API~\cite{troger2012distributed}), or OCCI
  (Open Cloud Computing Interface~\cite{edmonds2012toward}).
\item Data control: operations to manage data on the infrastructure,
  such as: upload, download, deletion, browsing, replication, caching,
  etc. Data movements can be triggered by the user in the science
  gateway (\texttt{$c_1$}), to upload input data or download processed
  data. They can also be performed by the workflow engine
  (\texttt{$c_2$}), to transfer workflow data (inputs, outputs, or
  intermediate results) across the infrastructure so that tasks can
  use it. The infrastructure might offer various data storage backends
  with heterogeneous interfaces. \closednote{Marc-Etienne}{this -c2- can be also trigerred
    by the user}{Tristan}{No, it can't: this is by definition an interaction between the workflow engine and the infrastructure. It is used to transfer data to/from tasks. I reworded a bit, is it clearer now?}
    {Marc-e}{I meant, "using c1", a user in some platforms (CBRAIN), can trigger
    some operations across the infrastructure, as in c2... not just up/download data.
    And I clearly see now that this is in fact the case in III, c is not differentiated. Closing this thread for good.}
  Tools and services such as JSAGA~\cite{reynaud2010uniform} or Data Avenue~\cite{hajnal2014data}
  can be used to homogeneize these interfaces.
\item Workflow control: operations to control workflow execution in an
  engine, including: workflow submission, monitoring, termination,
  etc. Workflow control can be coarse-grained (black box) or
  fine-grained (white box). In a coarse-grained model, the various
  tasks created by a workflow execution are masked and the user only
  has a global view of the workflow execution. In a fine-grained
  model, user is exposed to the workflow topology, i.e. to the outputs
  of the individual tasks, their statuses and so on.
\item Sub-task control: operations used by tasks to submit sub-tasks
  on the infrastructure, including: submission, monitoring,
  termination, deletion, etc. Sub-task control is similar to
  interaction \texttt{b}, except that information about the parent
  task, which submitted a sub-task, is usually available and used for
  additional control. For instance, the parent task may wait for all
  its sub-tasks to complete before finishing, and conversely all the
  sub-tasks may be terminated if the parent task is killed.
\item Pool-agent: specific to the pool architecture described in
  Section~\ref{sec:pool}. This is an interaction used when agents
  retrieve work from a central pool. It covers agent registration and
  de-registration to the pool, protocols to send work from the pool to
  the agent, mechanisms to update work status, and so on. A similar
  type of interaction is used in pilot-job
  systems~\cite{turilli2015comprehensive} and other agent-based
  computing models.
\item Workflow conversion: translation from one workflow language to
  another one. This interaction may not be available or implementable
  for every workflow language. It has been developed mostly for
  well-structured and relatively simple workflow language such as between GWorkflowDL
  and Scufl~\cite{OLAB-09}, and between the 4 languages that were involved in the
  SHIWA initiative. In SHIWA, workflow conversion is done through an
  intermediate representation, the Interoperable Workflow Intermediate
  Representation~\cite{plankensteiner-montagnat-etal:2011}, which
  allows to convert among $n$ workflow languages using $2n$
  interactions instead of $n^2$.
\end{enumerate}

@@

See Figure~\subref*{archi:tight}. The workflow engine is tightly
integrated with the science gateway, which means that it is deployed
on the same machine and potentially shares code, libraries and other
software components with the science gateway. For instance, the
workflow engine might be a portlet in a Liferay portal or a controller
in a Ruby on Rails application. The workflow engine and the science
gateway usually share a database where application, users and other
resources are stored\closednote{sorina}{VIP and Moteur also share the
  WorkflowsDB}{Tristan}{Yes, you are right. This is an example where
  such abstract architectures may not completely apply in real
  life. See comment about ``blended'' architectures in the
  discussion. In case of VIP, the sharing of the DB is more an
  implementation detail rather than a strong architectural pattern: we
  could avoid sharing the DB if we wanted.}.  In this model, task and
data control are both initiated from the science gateway. Interactions
\texttt{b} and \texttt{c$_2$} are initiated from the workflow engine
while \texttt{c$_1$} comes from other parts of the science gateway,
for instance a data management interface. As in any other model, the
installation of new workflows in the science gateway (\texttt{a$_2$})
and infrastructure (\texttt{a$_1$}) is done by an administrator, for
security reasons. This is the model adopted in the Catania Science
Gateway Framework~\cite{Ardizzone2012} (see specific documentation on
workflows\footnote{\url{http://bit.ly/1oQrzvQ}}), in the Distributed
application runtime environment
(DARE~\cite{maddineni2012distributed}), in
Galaxy~\cite{goecks2010galaxy}, and in LONI Pipeline
Environment~\cite{dinov2009efficient}. Note that tightly integrated
architectures may provide advanced workflow edition features which are
not covered by our analysis.
%The Catania Science Gateway Framework would be a good illustration here (workflow engine is a Java portlet), but I can't find a suitable figure about it.
%CIPRES~\cite{miller2010creating} (on which the Neuroscience gateway
%described in~\cite{sivagnanam2015early} is based) % \note{Tristan}{There is no workflow engine in there.}

@@

See Figure~\subref*{archi:import}. This is an abstract model
instantiated with the service invocation architecture for
consistency. Workflows are integrated in the science gateway through
format conversion from a native format to the science gateway
format. Workflow import is usually an offline process which is not
involved in workflow execution. Workflow import was implemented in the
SHIWA Simulation Platform through the IWIR language that provided a
common language for portability across grid workflow
systems~\cite{plankensteiner-prodan-etal:2013}.  Other systems rely on
workflow import. Tavaxy~\cite{Abouelhoda2012} allows to import, merge
and run Taverna~\cite{oinn2004taverna} and
Galaxy~\cite{goecks2010galaxy} workflows. When some workflow parts
cannot be imported, workflows are run by Tavaxy as nested workflows
using their native engine. The work in~\cite{delaGarza2016} describes
workflow import from KNIME~\cite{Berthold2008} to WS-PGRADE/gUSE and
from Galaxy~\cite{goecks2010galaxy} to WS-PGRADE/gUSE.
\closednote{Tristan}{Add description of a real system}{Tristan}{Would
  require to reproduce a figure from another paper. The concept seems
  simple enough to not require a figure.}

%Furthermore, both systems provide web-portals allowing users to share and publish their workflows: These are the myExperiment portal for Taverna [23, 24] and the Public Pages for Galaxy [13].

@@

\emph{Robustness of workflow execution} measures the likelihood that
workflow execution fails due to errors in the components or with the
interactions in the software architecture. Note that errors coming
from the infrastructure (e.g., unavailable data or terminated tasks) or
workflows (e.g., wrong user input or application errors) are not
covered since they do not stem directly from the software
architecture. Robustness is determined as the number of software
components and interactions that are specific to workflow execution,
which assumes that complex architectures tend to be more prone to failure:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Components (\texttt{R$_1$}): number of specific components
  involved in workflow executions. Science
gateway, for instance, is \emph{not} specific to workflow execution
since it is used to authenticate users, add new workflows, transfer
user data, etc.
\item Interactions (\texttt{R$_2$}): number of specific interactions
  involved in workflow executions.
\end{itemize}
These specific components and interactions are represented in red in
the architecture diagrams in
Figure~\ref{fig:architectures}. \texttt{R$_1$} and \texttt{R$_2$}  are summed to obtain the total score for this
criterion, which measures the total number of software pieces that are
specifically involved in the workflow execution.

\emph{Extensibility} measures the difficulty to replace or add
elements in the architecture. It is determined as the number of
interactions and components that need to be modified when a new
element is added. Modification of a component is required when its
code needs to be updated or recompiled (science gateway or workflow
service), or when a new piece of software has to be installed
(infrastructure only). Modification of an interaction is deemed
necessary when the parameters involved in this interaction are
modified.  Extensibility breaks down in 4 metrics depending on the
type of element that has to be added or replaced:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item New engine type (\texttt{E$_1$}): number of interactions or
  components to modify to integrate a new type of workflow engine in
  the architecture. Workflow engines belong to different types when
  they cannot be invoked using the same interface. Adding a new type
  of workflow engine allows to execute more workflows in the science
  gateway.
\item New engine version (\texttt{E$_2$}): number of interactions or
  components to modify to integrate a new version of a workflow engine
  in the architecture, assuming that another version of the same
  engine type is already available. Different versions of a workflow
  engine share the same interface, i.e. they can be invoked using the
  same software. When this is not the case, the different versions are
  considered as different engine types.
\item New workflow (\texttt{E$_3$}): adding a new workflow is a very
  common operation that does not require modifying software components
  or interactions. We measure the difficulty to integrate a new
  workflow by counting the number of interactions required
  to integrate a new workflow in the architecture, assuming
  that the engine type and version required to execute this workflow
  are already available.
\item New infrastructure (\texttt{E$_4$}): number of interactions or
  components to modify to integrate a new type of infrastructure in
  the architecture. Adding a new infrastructure allows to provide more
  computing or storage power, to access specific types of resources
  (e.g., GPUs, clouds), or to enforce execution policies (e.g.,
  constrain data to remain in a particular network domain). 
%\item New science gateway: \todo{Not sure if this makes sense}.
\end{itemize}
The 4 metrics are summed to obtain a global index that measures the
difficulty to extend the architecture. Note that some extensions may
involve several metrics in practice. For instance, adding a new type
of engine may help integrate new infrastructures when interactions
\texttt{b} and \texttt{c$_2$} are already present for the new engine
and infrastructure.  \closednote{Rafael}{Adding a new engine type may
  also allow run in different infrastructures. Maybe it is worth to
  put a sentence to clarify that the metric account for it
  separately.}{Tristan}{It will not *allow* because interaction c1
  would still need to be developed. But it will definitely facilitate
  it. I added a note on this after the description of the metrics.}

>>>>

\definecolor{ao(english)}{rgb}{0.0, 0.5, 0.0}
\newcommand{\todo}[2]{\pdfmargincomment[color=red,author=#1,open=true]{#2}}
\newcommand{\note}[2]{\pdfmargincomment[color=yellow,author=#1,open=true]{#2}}
\newcommand{\closednote}[4]{\pdfmargincomment[color=gray,author=#1-#3,open=true]{#1: #2 \textCR \textCR #3: #4}}

@@

\note{Marc-e}{Add Samir Das?}

@@

Architectures showing reduced system complexity will greatly
facilitate the implementation effort towards robust interactions
which, in turn, have a positive impact on characteristics such as
gateway predictability, transparence, general reliability and results
traceability and reproducibility; characteristics that are also key to
a good user experience. \closednote{Marc-Etienne}{I find that
  something is weak in this last paragraph. Robustness will help with
  transparency, sure, but it seems only a small facet.  Also, the text
  seem to timidly favor architectures with less
  components/interactions without much explanation... Complex systems
  can be robust, but at greater costs...  Here is one: "Architectures
  showing reduced system complexity will greatly facilitate the
  implementation effort towards robust interactions which, in turn,
  have a positive impact on characteristics such as gateway
  predictability, transparence, general reliability and results
  traceability and reproducibility; characteristics that are also key
  to a good user experience."}{Tristan}{Agreed, fixed as suggested.}

@@

Various science gateways have been developed, including frameworks
used as building blocks to assemble science gateways. The main
frameworks are Apache Airavata~\cite{marru2011apache}, the Catania
Science Gateway Framework~\cite{Ardizzone2012} and
WS-PGRADE/gUSE~\cite{Kacsuk2012}. Numerous science gateway instances
were built using such frameworks\note{Tristan}{Add references}. Other
gateway instances were also built independently from any framework,
using regular web development tools and libraries to access
distributed systems. CBRAIN and VIP are two such science gateways,
from which we gathered experience about the architectures described in
this paper.

@@

\begin{figure*}
\centering
\hspace*{0.2\columnwidth}
\subfloat[Tight integration]{
\def\svgwidth{0.6\columnwidth}
\input{figures/tight.pdf_tex}
\label{archi:tight}
}
\hfill
\subfloat[Service invocation]{
\def\svgwidth{0.6\columnwidth}
\input{figures/service.pdf_tex}
\label{archi:service}
\hspace*{0.2\columnwidth}
}\\
\hspace*{0.2\columnwidth}
\subfloat[Sub-tasking]{
\def\svgwidth{0.6\columnwidth}
\input{figures/sub-task.pdf_tex}
\label{archi:sub-task}
}
\hfill
\subfloat[Pool]{
\def\svgwidth{0.6\columnwidth}
\input{figures/agent.pdf_tex}
\label{archi:agent}
\hspace*{0.2\columnwidth}
}\\
\subfloat[Nested workflows. Left: abstract model. Right: instantiation with service invocation.]{
\def\svgwidth{0.9\columnwidth}
\input{figures/nested-1.pdf_tex}
\def\svgwidth{0.9\columnwidth}
\input{figures/nested-2.pdf_tex}
\label{archi:nested}
}\\
\subfloat[Workflow import. Left: abstract model. Right: instantiation with service invocation.]{
\def\svgwidth{0.5\columnwidth}
\input{figures/import-1.pdf_tex}
\hfill
\def\svgwidth{0.5\columnwidth}
\input{figures/import-2.pdf_tex}
\label{archi:import}
}
\caption{Architectures}
\label{fig:architectures}
\end{figure*}
\begin{table*}
\centering
\footnotesize
\rowcolors{1}{white}{gray!25}
\begin{tabular}{ll}
  \textbf{Architecture} & \textbf{Systems} \\
  \hline
  Tight & \pbox{1.5\columnwidth}{
          Catania Science Gateway Framework~\cite{Ardizzone2012},
          Distributed application runtime environment (DARE~\cite{maddineni2012distributed}),
          DECIDE~\cite{ardizzone2012decide}$^{c,n}$, LONI Pipeline Environment$^n$~\cite{dinov2009efficient}
          }
  \\
  Service & \pbox{1.5\columnwidth}{
            Apache Airvata~\cite{marru2011apache}, e-BioInfra~\cite{shahand2015data}$^{w,n}$, HubZero with Pegasus~\cite{CPE:CPE3257}, MoSGrid~\cite{krüger2014mosgrid}$^w$, System in~\cite{wu2010accelerating}, Vine Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual Imaging Platform~\cite{GLAT-13}$^n$, WS-PGRADE/gUSE framework~\cite{Kacsuk2012},
            Science gateways in ~\cite{kacsuk2014science}$^w$
            } \\
  Sub-task & CBRAIN and PSOM~\cite{GLAT-16}$^n$, CBRAIN and FSL$^n$\\
  Pool & SHIWA pool~\cite{ROGE-13}\\
  Nested & \pbox{1.5\columnwidth}{SHIWA Simulation Platform (Coarse-Grained Interoperability~\cite{terstyanszky2014enabling})$^w$, HubZero with Pegasus (via hierarchical workflows)~\cite{Deelman201517} \todo{Tristan}{Add others}
  }\\
  Import & SHIWA Simulation Platform (Fine-Grained Interoperability)~\cite{plankensteiner-prodan-etal:2013}$^w$
\end{tabular}
\caption{Classification of science gateways based on the architecture used to integrate workflow engines. $^c$: based on the Catania Science Gateway Framework. $^w$: based on WS-PGRADE/gUSE. $^n$: used for neuroimaging.}
\label{table:system-classification}
\end{table*}

@@

\item Workflow integration: consists in adding a new workflow to the
  system so that users can execute it. It is triggered by an
  administrator of the science gateway and it results in an interface,
  for instance a web form, where users can enter the parameters of the
  workflow to be executed. The interaction has two aspects:
  (\texttt{a$_1$}) the programs used in the workflow are installed on
  the infrastructure, which may require administrative privileges on
  the infrastructure; (\texttt{a$_2$}) the workflow is configured in
  the science gateway so that it becomes available to users. Note that
  integrating a workflow is not the same process as integrating a
  workflow \emph{engine}.

\item Task control: operations to manage tasks on the infrastructure,
  including: authentication, submission, monitoring, termination,
  deletion, etc. Controlling tasks requires to deal with the
  heterogeneous batch managers and meta-schedulers that might be
  available on the infrastructure. When the infrastructure is a grid
  or a cloud, it may for instance be achieved using libraries that
  implement standards such as SAGA (Simple API for Grid
  Applications~\cite{goodale2006saga}), DRMAA (Distributed Resource
  Management Application API~\cite{troger2012distributed}), or OCCI
  (Open Cloud Computing Interface~\cite{edmonds2012toward}).
\item Data control: operations to manage data on the infrastructure,
  such as: upload, download, deletion, browsing, replication, caching,
  etc. Data movements can be triggered by the user in the science
  gateway (\texttt{$c_1$}), to upload input data or download processed
  data. They can also be performed by the workflow engine
  (\texttt{$c_2$}), to transfer workflow data (inputs, outputs, or
  intermediate results) across the infrastructure so that tasks can
  use it. The infrastructure might offer various data storage backends
  with heterogeneous interfaces. \closednote{Marc-Etienne}{this -c2- can be also trigerred
    by the user}{Tristan}{No, it can't: this is by definition an interaction between the workflow engine and the infrastructure. It is used to transfer data to/from tasks. I reworded a bit, is it clearer now?} Tools and services such as
  JSAGA~\cite{reynaud2010uniform} or Data Avenue~\cite{hajnal2014data}
  can be used to homogeneize these interfaces.
\item Workflow control: operations to control workflow execution in an
  engine, including: workflow submission, monitoring, termination,
  etc. Workflow control can be coarse-grained (black box) or
  fine-grained (white box). In a coarse-grained model, the various
  tasks created by a workflow execution are masked and the user only
  has a global view of the workflow execution. In a fine-grained
  model, user is exposed to the workflow topology, i.e. to the outputs
  of the individual tasks, their statuses and so on.
\item Sub-task control: operations used by tasks to submit sub-tasks
  on the infrastructure, including: submission, monitoring,
  termination, deletion, etc. Sub-task control is similar to
  interaction \texttt{b}, except that information about the parent
  task, which submitted a sub-task, is usually available and used for
  additional control. For instance, the parent task may wait for all
  its sub-tasks to complete before finishing, and conversely all the
  sub-tasks may be terminated if the parent task is killed.
\item Pool-agent: specific to the pool architecture described in
  Section~\ref{sec:pool}. This is an interaction used when agents
  retrieve work from a central pool. It covers agent registration and
  de-registration to the pool, protocols to send work from the pool to
  the agent, mechanisms to update work status, and so on. A similar
  type of interaction is used in pilot-job
  systems~\cite{turilli2015comprehensive} and other agent-based
  computing models.
\item Workflow conversion: translation from one workflow language to
  another one. This interaction may not be available or implementable
  for every workflow language. It has been developed mostly for
  well-structured and relatively simple workflow language such as between GWorkflowDL
  and Scufl~\cite{OLAB-09}, and between the 4 languages that were involved in the
  SHIWA initiative. In SHIWA, workflow conversion is done through an
  intermediate representation, the Interoperable Workflow Intermediate
  Representation~\cite{plankensteiner-montagnat-etal:2011}, which
  allows to convert among $n$ workflow languages using $2n$
  interactions instead of $n^2$.
\end{enumerate}

@@

See Figure~\subref*{archi:tight}. The workflow engine is tightly
integrated with the science gateway, which means that it is deployed
on the same machine and potentially shares code, libraries and other
software components with the science gateway. For instance, the
workflow engine might be a portlet in a Liferay portal or a controller
in a Ruby on Rails application. The workflow engine and the science
gateway usually share a database where application, users and other
resources are stored\closednote{sorina}{VIP and Moteur also share the
  WorkflowsDB}{Tristan}{Yes, you are right. This is an example where
  such abstract architectures may not completely apply in real
  life. See comment about ``blended'' architectures in the
  discussion. In case of VIP, the sharing of the DB is more an
  implementation detail rather than a strong architectural pattern: we
  could avoid sharing the DB if we wanted.}.  In this model, task and
data control are both initiated from the science gateway. Interactions
\texttt{b} and \texttt{c$_2$} are initiated from the workflow engine
while \texttt{c$_1$} comes from other parts of the science gateway,
for instance a data management interface. As in any other model, the
installation of new workflows in the science gateway (\texttt{a$_2$})
and infrastructure (\texttt{a$_1$}) is done by an administrator, for
security reasons. This is the model adopted in the Catania Science
Gateway Framework~\cite{Ardizzone2012} (see specific documentation on
workflows\footnote{\url{http://bit.ly/1oQrzvQ}}), in the Distributed
application runtime environment (DARE~\cite{maddineni2012distributed})
and in LONI Pipeline Environment~\cite{dinov2009efficient}.
%The Catania Science Gateway Framework would be a good illustration here (workflow engine is a Java portlet), but I can't find a suitable figure about it.
%CIPRES~\cite{miller2010creating} (on which the Neuroscience gateway
%described in~\cite{sivagnanam2015early} is based) % \note{Tristan}{There is no workflow engine in there.}

@@

See Figure~\subref*{archi:import}. This is an abstract model
instantiated with the service invocation architecture for
consistency. Workflows are integrated in the science gateway through
format conversion from a native format to the science gateway
format. Workflow import is usually an offline process which is not
involved in workflow execution. Workflow import was implemented in the SHIWA
Simulation Platform through the IWIR language that provided a common
language for portability across grid workflow
systems~\cite{plankensteiner-prodan-etal:2013}.

\todo{Tristan}{Add description of a real system}

\todo{Tristan}{Refer to ``From the desktop to the grid: scalable bioinformatics via workflow conversion Luis de la GarzaEmail author, Johannes Veit, Andras Szolek, Marc Röttig, Stephan Aiche, Sandra Gesing, Knut Reinert and Oliver Kohlbacher''}

@@

\emph{Robustness of workflow execution} measures the likelihood that
workflow execution fails due to errors in the components or with the
interactions in the software architecture. Note that errors coming
from the infrastructure (e.g., unavailable data or terminated tasks) or
workflows (e.g., wrong user input or application errors) are not
covered since they do not stem directly from the software
architecture. Robustness is determined as the number of software
components and interactions that are specific to workflow execution,
which assumes that complex architectures tend to be prone to failure:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Components (\texttt{R$_1$}): number of specific components
  involved in workflow executions. Science
gateway, for instance, is \emph{not} specific to workflow execution
since it is used to authenticate users, add new workflows, transfer
user data, etc.
\item Interactions (\texttt{R$_2$}): number of specific interactions
  involved in workflow executions.
\end{itemize}
These specific components and interactions are represented in red in
the architecture diagrams in
Figure~\ref{fig:architectures}. \texttt{R$_1$} and \texttt{R$_2$}  are summed to obtain the total score for this
criterion, which measures the total number of software pieces that are
specifically involved in the workflow execution.

\emph{Extensibility} measures the difficulty to replace or add
elements in the architecture. It is determined as the number of
interactions and components that need to be modified when a new
element is added. Modification of a component is required when its
code needs to be updated or recompiled (science gateway or workflow
service), or when a new piece of software has to be installed
(infrastructure only). Modification of an interaction is deemed
necessary when the parameters involved in this interaction are
modified.  Extensibility breaks down in 4 metrics depending on the
type of element that has to be added or replaced:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item New engine type (\texttt{E$_1$}): number of interactions or
  components to modify to integrate a new type of workflow engine in
  the architecture. Workflow engines belong to different types when
  they cannot be invoked using the same interface. Adding a new type
  of workflow engine allows to execute more workflows in the science
  gateway.
\item New engine version (\texttt{E$_2$}): number of interactions or
  components to modify to integrate a new version of a workflow engine
  in the architecture, assuming that another version of the same
  engine type is already available. Different versions of a workflow
  engine share the same interface, i.e. they can be invoked using the
  same software. When this is not the case, the different versions are
  considered as different engine types.
\item New workflow (\texttt{E$_3$}): adding a new workflow is a very
  common operation that does not require modifying software components
  or interactions. We measure the difficulty to integrate a new
  workflow by counting the number of interactions required
  to integrate a new workflow in the architecture, assuming
  that the engine type and version required to execute this workflow
  are already available.
\item New infrastructure (\texttt{E$_4$}): number of interactions or components to
  modify to integrate a new type of infrastructure in the
  architecture. Adding a new infrastructure allows to provide more
  computing or storage power, to access specific types of resources
  (e.g., GPUs, clouds), or to enforce execution policies
  (e.g., constrain data to remain in a particular network domain).\note{Rafael}{Adding a new engine type may also allow run in different infrastructures. Maybe it is worth to put a sentence to clarify that the metric account for it separately.}
%\item New science gateway: \todo{Not sure if this makes sense}.
\end{itemize}
The 4 metrics are summed to obtain a global index that measures the
difficulty to extend the architecture.

\emph{Scalability} corresponds to the ability of the architecture to
cope with high workloads. It is measured by counting the potential
scalability issues in the architectures, i.e. the lack of scalability
features. Features are evaluated using a 3-level metric: 0 means that
the feature is very easy to enable, 1 means that it can be implemented
but with some difficulty, and 2 means that the feature could only be
implemented with a nonsensical amount of effort. Four different
features are identified:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Multiple engine instances (\texttt{S$_1$}): possibility to have
  more than 1 engine instance in the architecture. Workflow engines
  may require important amounts of resources when several workflows,
  or large workflows, are executed. At some point, it may be required
  for the science gateway to distribute the load among several
  engines. \texttt{S$_1$}=0 when adding a new engine instance is a
  fully automated process, i.e. the workflow engine only has to be
  started. In this case, elastic engines are possible, i.e. some kind
  of auto-scaling mechanism can be implemented to control the number
  of engine instances in the architecture. \texttt{S$_1$}=1 when
  adding a new engine instance requires some form of manual
  intervention (e.g., two-factor authentication), which prevents easy
  implementations of elastic engines. \texttt{S$_1$}=2 when new engine
  instances cannot be added.
\item Distributed engines (\texttt{S$_2$}): possibility to distribute
  the execution \emph{of a single workflow} among different engine
  instances. In our scope, this feature focuses on the capabilities of
  the architecture rather than these of the workflow
  engine. \texttt{S$_2$}=0 when distributed engines are possible in
  the architecture, and \texttt{S$_2$}=1 when they require specific
  developments in the workflow engine.\note{Rafael}{$S_2$ would be 2 if the paradigm does not allow to distribute the orchestration? e.g., workflow executions implemented within MPI jobs, in Pegasus PMC and dispel4Py.}
\item Task scheduling (\texttt{S$_3$}): task scheduling is a difficult
  issue that depends more on the implementation of specific algorithms
  in the science gateway, workflow engine and infrastructure than on
  the architecture used to integrate the workflow engine in the
  science gateway. Some architectures, however, complexify the task
  scheduling problem by introducing additional software layers or
  create tasks with specific characteristics. \texttt{S$_3$}=0 when
  the architecture does not introduce any additional complexity to the
  scheduling problem, and \texttt{S$_3$}=1 otherwise.
\end{itemize}
The 3 metrics are summed to obtain a global measure of the scalability
issues that are present in the architecture.
